### 2014-07-25 Studies/2012_Chicago/Report/Chicago_Fire.tex
Reports/Docs: Change font size to 12 pt
Chicago Report: Make edits throughout report

\documentclass[11pt,oneside]{book}

@@

@@

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements and standards, improve the understanding of the behavior, prevention and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) to provide insight on the fire development and thermal conditions of other line of duty death (LODD) fires~\cite{Madrzykowski:1,Iowa,Texas,Cook_County,Grosshandler:Station,Bryner:Charleston,barowy:texas}. The overriding objective of all of these studies is to improve the safety of fire fighters.

On November 2, 2012, a fire in a 2 1/2 story wood frame residential structure claimed the life of a Chicago firefighter. NIST examined the fire dynamics of this incident at the request of NIOSH and Chicago Fire Department (CFD). Computer simulations of the fire using Fire Dynamics Simulator and Smokeview, provide insight on the fire development and thermal conditions that may have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine with physics-based calculations the effect of flow path on the temperature, pressure, and fire conditions in this multi-story residential structure.
\item To provide visualization of fire behavior representative of what Chicago Fire Department members experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For analysis purposes, details regarding times are considered approximate values. The following incident narrative is taken from that NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' \footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} alarm for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure?s attic. Per fire department standard operating procedures, dispatch initiated a RIT \footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.} response. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2nd floor. He noticed only a light haze throughout and glow around the Sector 3 door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a horseshoe load which is 100 feet of 2-$\tfrac{1}{2}$ inch, a gated wye, and 150 feet of 1-$\tfrac{3}{4}$ inch hoseline to go to the 2nd floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1st floor apartment and worked his way to [the rear of the structure]. He opened the back door to the covered porch and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2nd floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2-$\tfrac{1}{2}$ inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [on the left side]. A TL39 firefighter went to [rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door to the first floor apartment, preceded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgement from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2nd floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the victim and other crew member untangled. The TL39 Lieutenant and SQ5 fire fighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

At 1727 hours, the TL39 crew had just completed the first hole in the roof about midway on [the right side] roof with minimal fire showing, when they heard the Mayday. The IC verified that a firefighter was down, called a Mayday, and requested a 2-11 Assignment \footnote{A 2-11 alarm is a 2nd alarm upgrade. The response is four engines, two trucks, one tower ladder, two battalion chiefs, one deputy district chief, one air mask service bottle truck, and a media affairs unit.}. Dispatch initiated a 2-11 alarm. SQ5 and other members on the 2nd floor grabbed the [E123 captain] and got him down the stairs. The TL39 crew, with assistance from a T33 and SQ5 firefighter, had just completed a second hole on [right side] of the roof about a third of the way back when conditions worsened. At 1729 hours, the roof ventilation crew was back in the aerial basket when they noticed the victim being brought out to the front yard. The victim was nonresponsive in the front yard and CPR was successfully performed. The IC met the A19 crew and escorted them to the [E123 captain]. The revived [captain] was responsive and talking to the paramedics as he was loaded into the ambulance. At 1738 hours, the [captain] was transported to the local hospital where he had complications during airway management and died.
\end{quote}

\noindent For an overview of timeline of events, see Table~\ref{tab:fire_info}. A plan view of the structure and location where the captain collapsed can be found in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NISOH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{l l}
\toprule[1.5pt]
Incident Time [hr:min:s] & Fire Behavior / Fireground Operation  \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm $-$ two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on 2nd floor with 1-$\tfrac{3}{4}$ inch hoseline. \\[.25cm]
17:21:30 & 1st floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123: captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
FDS is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is given in reference~\cite{FDS_Math_Guide}.

Smokeview is a software tool designed to visualize the results of the FDS~\cite{Smokeview_Users_Guide}.  Smokeview visualizes smoke and other attributes of the modeled fire using traditional scientific methods such as displaying tracer particle flow, two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data such as temperature and flow vectors showing flow direction and magnitude.  Smokeview allows the fire and fire movement to be visualized.  This is done by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS.  Smokeview also visualizes static data at particular times using 2D or 3D contours of data such as temperature and flow vectors showing flow direction and magnitude.

Input data from a number of sources must be collected and documented in order to simulate a fire with FDS or any other model. For the model presented here, information was obtained from two main sources. From the fire scene the following information was gathered: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. From witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, order and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire. The following subsections will describe the inputs used to construct the simulation, including any engineering approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure was a 2 1/2 story residential building with balloon frame construction and gable roof. \footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction, is that without fire stops, there is void space for fire to potentially travel between the floors.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint (cf. Fig.~(\ref{fig:simp_geom})). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). Interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally there was a half-story attic that was used to house the structure's hot water heater. There were three sets of stairs in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, see Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on interior conditions can be found in Section~\ref{results}. Fig.~(\ref{fig:alpha_ex}) shows the post-incident exterior of the front and rear of the structure. Note that the majority of the damage is to the structure's attic and second floor rear porch. Fully dimensioned drawings of the first and second floor interiors can be found in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

\begin{figure}[h!]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=\textwidth]{../Figures/1st_Floor_Metric} \\
\includegraphics[width=\textwidth]{../Figures/2nd_Floor_Metric}
\end{tabular*}
\caption{Plan view of 1st (top) and 2nd (bottom) floor of structure. Stairs and key rooms are identified Drawn using measurements collected by NIST staff.}
\label{fig:simp_geom}
\end{figure}

\begin{figure}[h!]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photograph showing the post-incident front (left) and rear (right) exterior of the structure.}
\label{fig:alpha_ex}
\end{figure}
%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[h!]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

To estimate the fuel, fire size, and structure ventilation, on-scene videos and post-incident reports are used. Based on these findings, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch consisted of a fuel load that is mainly wood: the rear porch was enclosed with wood planks exteriorly covered by foam board and vinyl siding, the roof was asphalt shingles over wood planking, and the attic and porch had oriented strand board (OSB) floors.

Wood can be represented by the the following chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; {\mathrm{kg_{CO}}/\mathrm{kg_{wood}}}$) and soot ($y_{\mathrm{C}}=0.015 \; {\mathrm{kg_{C}}/\mathrm{kg_{wood}}}$) found from the SFPE Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood oxidation can written:

@@

The heat of combustion of wood used in this simulation is 16400 kJ/kg, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the energy release per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

&REAC ID = 'wood' 
    FUEL = 'WOOD', 
    HEAT_OF_COMBUSTION=16400,
    SOOT_YIELD = 0.015,
    CO_YIELD = 0.004/
\end{lstlisting}

Note that using the input lines above will invoke the simple chemistry reaction balance in FDS. This means that fuel and air react to form only CO$_2$, CO, H$_2$O, Soot, and N$_2$. If there is a desire for any other combustion products then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on above input lines FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism  states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users to the option to use a more complex finite-rate combustion mechanism, we do not have sufficient evidence to justify deviating from the default specifications.

To estimate the energy release per unit time and unit area (kW/m$^2$) of wood, we use work by Babrauskas and Grayson~\cite{babrauskas1990}. Experiments were conducted in a cone calorimeter \footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials.} to determine the 5 minute average of HRRPUA  \footnote{HRRPUA is the heat release rate per unit area.} (kW/m$^2$) for several different types of wood over a range of radiant heat fluxes (kW/m$^2$). Babrauskas and Grayson~\cite{babrauskas1990} suggest that the heat release rate per unit area for a variety of woods is approximately 50~kW/m$^2$.

>>>>

\documentclass[12pt,oneside]{book}

@@

% Extra packages
\usepackage{xfrac}

@@

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Cook_County,Grosshandler:Station,Bryner:Charleston,barowy:texas}. The objective of all of these studies has been to improve the safety of fire fighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago firefighter. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using Fire Dynamics Simulator and Smokeview to provide insight into the fire development and thermal conditions that might have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding times are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the Sector 3 door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a horseshoe load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the back door to the covered porch and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door to the first floor apartment, preceded to walk in, saw no fire then backed out.

At 1723 hours, the IC radioed the [E123 captain] that there was heavy fire in the covered porch and attic area and that E49 was going to put water on the fire, around the [rear] attic window, but there was no acknowledgment from the [captain]. E49 proceeded to put water on the [rear of the structure]. The IC returned out front to the command post and donned his turnout gear. The lieutenant from TL39, the lieutenant and 2 firefighters from SQ5, and a E123 firefighter/paramedic (FF/PM) were near the kitchen area on the 2\textsuperscript{nd} floor when they heard a loud commotion. The FF/PM heard the [E123 captain] yell ``get out of here.'' The FF/PM had no radio and he couldn't locate the [captain's] radio so he yelled Mayday as he tried to get the [captain] and other crew member untangled. The TL39 Lieutenant and SQ5 fire fighters heard the FF/PM's verbal Mayday and the SQ5 Lieutenant tried to transmit a Mayday over heavy radio traffic.

At 1727 hours, the TL39 crew had just completed the first hole in the roof about midway on [the right side] roof with minimal fire showing, when they heard the Mayday. The IC verified that a firefighter was down, called a Mayday, and requested a 2-11 Assignment\footnote{The department defines a 2-11 alarm as a second alarm upgrade. The response is four engines, two trucks, one tower ladder, two battalion chiefs, one deputy district chief, one air mask service bottle truck, and a media affairs unit.}. Dispatch initiated a 2-11 alarm. SQ5 and other members on the 2\textsuperscript{nd} floor grabbed the [E123 captain] and got him down the stairs. The TL39 crew, with assistance from a T33 and SQ5 firefighter, had just completed a second hole on [the right side] of the roof about a third of the way back when conditions worsened. At 1729 hours, the roof ventilation crew was back in the aerial basket when they noticed the [captain] being brought out to the front yard. The [captain] was nonresponsive in the front yard and CPR was successfully performed. The IC met the A19 crew and escorted them to the [E123 captain]. The revived [captain] was responsive and talking to the paramedics as he was loaded into the ambulance. At 1738 hours, the [captain] was transported to the local hospital where he had complications during airway management and died.
\end{quote}

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NISOH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
(hh:mm:ss)     &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on 2nd floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & 1st floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

Smokeview is a software tool designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}.  Smokeview visualizes smoke and other attributes of the fire simulation using traditional scientific methods such as displaying tracer particle flow, two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data such as temperature and flow vectors showing flow direction and magnitude.  Smokeview allows the fire and fire movement to be visualized.  This is done by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS.  Smokeview also visualizes static data at particular times using 2D or 3D contours of data such as temperature and flow vectors showing flow direction and magnitude.

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, the sequence and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire. The following subsections will describe the inputs that were used to construct the simulation, including any engineering approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~(\ref{fig:simp_geom}). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~(\ref{fig:alpha_ex}) shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

\begin{figure}[h!]
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
\includegraphics[width=\textwidth]{../Figures/1st_Floor_Metric} \\
\includegraphics[width=\textwidth]{../Figures/2nd_Floor_Metric}
\end{tabular*}
\caption[Plan view of 1st (top) and 2nd (bottom) floor of structure.]{Plan view of 1st (top) and 2nd (bottom) floor of structure. Stairs and key rooms are identified using measurements collected by NIST.}
\label{fig:simp_geom}
\end{figure}

\begin{figure}[h!]
\centering
\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
   \scalebox{1.0}{ \includegraphics[width=3in]{../Figures/exterior_alpha} } &
   \scalebox{1.0}{ \includegraphics[width=2.67in]{../Figures/exterior_charlie} } \\
\end{tabular*}
\caption{Photographs showing the front (left) and rear (right) exterior of the structure after the incident.}
\label{fig:alpha_ex}
\end{figure}
%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%\begin{figure}[h!]
%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%\end{tabular*}
%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%\label{fig:smv_interior}
%\end{figure}

@@

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these findings, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board (OSB).

Wood can be represented by the the following chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; {\mathrm{kg_{CO}}/\mathrm{kg_{wood}}}$) and soot ($y_{\mathrm{C}}=0.015 \; {\mathrm{kg_{C}}/\mathrm{kg_{wood}}}$) found from the SFPE Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood oxidation can written as:

@@

The value for the heat of combustion of wood used in this simulation was 16,400 kJ/kg, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion quantifies the amount of energy per unit mass of the fuel. The following text defines Eq.~\ref{eq:wood_comb} for an FDS input file with the desired fuel and reaction properties discussed above:

@@

&REAC ID = 'wood' 
    FUEL = 'WOOD', 
    HEAT_OF_COMBUSTION = 16400,
    SOOT_YIELD = 0.015,
    CO_YIELD = 0.004/
\end{lstlisting}

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion for other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users to the option to use a more complex finite-rate combustion mechanism, we do not have sufficient evidence to justify deviating from the default specifications.

To estimate the energy release rate per unit area (kW/m$^2$) of wood, we use work by Babrauskas and Grayson~\cite{babrauskas1990}. Experiments were conducted in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials.} to determine the 5 minute average of HRRPUA\footnote{HRRPUA is the heat release rate per unit area.} (kW/m$^2$) for several different types of wood over a range of radiant heat fluxes (kW/m$^2$). Babrauskas and Grayson~\cite{babrauskas1990} suggest that the heat release rate per unit area for a variety of woods is approximately 50~kW/m$^2$.

To calculate the total fire size, the appropriate burning area (exposed wood area) of the attic and second floor porch enclosure needs to be determined. Based on the floor plan of the structure, the underside of the roof is approximately 134~m$^2$ (1453~ft$^2$) with 7~m$^2$ (75~ft$^2$) walls on the front and rear sides. Additionally, there is 105~m$^2$ (1130~ft$^2$) of floor area in the attic. Examination of post-incident images indicated that the floor immediately over the rear porch burned away. That area was considered to be part of the porch fire. Based on post-incident damage, approximately 50~\% of the remaining floor area and 90~\% of the roof was considered to be fuel for the attic fire. To estimate the total attic fire size, the underside roof area (120~m$^2$), two end walls (14~m$^2$), and the floor area (43~m$^2$ or 463~ft$^2$) were included. Applying an HRRPUA of wood of 50~kW/m$^2$ results in a potential peak fire size of 9~MW. To estimate the size of the porch fire, the four walls of the porch enclosure (42~m$^2$ or 452~ft$^2$) and the floor and ceiling of the porch (24~m$^2$ or 258 ft$^2$) were included, which results in a total porch area of 66~m$^2$ (689~ft$^2$). Using the same HRRPUA of wood of 50~kW/m$^2$, the potential peak fire size on the porch is approximately 3.3~MW. Therefore, the maximum fire size for this structure was estimated to be 12.3~MW. Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The first assumption is that the attic fire was specified to start at the beginning of the simulation and increase to a value of 9~MW in 10~s. The porch fire was specified to start 40~s into the simulation and increase to a value of 3.3~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations (2 sources for the attic, 1 source for the porch). The following lines define the source fires in the FDS input file:

\begin{lstlisting}
&SURF ID='FIRE', HRRPUA=2250., COLOR='ORANGE', TAU_Q=10 /
&VENT XB=-13.5,-13,1,5,6.16,6.16, SURF_ID='FIRE' / 2 m^2
&VENT XB=-14.5,-14,1,5,6.16,6.16, SURF_ID='FIRE' / 2 m^2
&VENT XB=-16.5,-15.5,1,2.5,3.14,3.14, SURF_ID='FIRE', DEVC_ID='SECOND FIRE'/ 1.5 m^2
&DEVC XYZ = -17.2,2,0 ID = 'SECOND FIRE', QUANTITY = 'TIME', SETPOINT = 40, INITIAL_STATE = .FALSE. /
\end{lstlisting}
Note that the coordinates here are unique for the vents are specific to the input files used for these simulations. The key point is that they are 2-D planes located at a specific area.
